---
title: 'Reproducible Research: Peer Assessment 1'
author: "genesysx"
date: "6 January 2016"
        output: html_document
                keep_md: true
---

# Reproducible Research: Peer Assessment 1
library(ggplot2)
library(data.table)

## Loading and preprocessing the data
##### 1. Load the data (i.e. read.csv())
```{r, results='markup', warning=TRUE, message=TRUE}

activityData <- read.csv('activity.csv')
```
##### 2. Process/transform the data (if necessary) into a format suitable for your analysis
```{r}
#activityData$interval <- strptime(gsub("([0-9]{1,2})([0-9]{2})", "\\1:\\2", activityData$interval), format='%H:%M')
```

-----

## What is mean total number of steps taken per day?
```{r}
stepsPerDay <- tapply(activityData$steps, activityData$date, sum, na.rm=TRUE)
```

##### 1. Make a histogram of the total number of steps taken each day
```{r, echo= TRUE}
qplot(stepsPerDay, xlab='Total Steps/ Day', ylab='Freq Using BinWidth 500', binwidth=500, geom="histogram", fill=I("blue"))

```

##### 2. Calculate and report the mean and median total number of steps taken per day
```{r}
stepsPerDayMean <- mean(stepsPerDay)
stepsPerDayMedian <- median(stepsPerDay)
```
* Mean: `r stepsPerDayMean`
* Median:  `r stepsPerDayMedian`

-----

## What is the average daily activity pattern?
```{r}
averageStepsPerTimeBlock <- aggregate(x=list(meanSteps=activityData$steps), by=list(interval=activityData$interval), FUN=mean, na.rm=TRUE)
```

##### 1. Make a time series plot
```{r}
ggplot(data=averageStepsPerTimeBlock, aes(x=interval, y=meanSteps)) +
    geom_line() +
    xlab("5 Minute Interval") +
    ylab("Average Number of Steps Taken") 
```

##### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
mostSteps <- which.max(averageStepsPerTimeBlock$meanSteps)
ans<-averageStepsPerTimeBlock[mostSteps,'interval']
timeMostSteps<-gsub("[.]", "\\1:", ans/100)

```

* Most Steps at: `r timeMostSteps`

----

## Imputing missing values
##### 1. Calculate and report the total number of missing values in the dataset 
```{r}
numMissingValues <- length(which(is.na(activityData$steps)))
```

* Number of missing values: `r numMissingValues`

##### 2. Devise a strategy for filling in all of the missing values in the dataset.
##### 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
DT <- data.table(activityData)
setkey(DT, interval)
DT[,steps := ifelse(is.na(steps), median(steps, na.rm=TRUE), steps), by=interval]
```

##### 4. Make a histogram of the total number of steps taken each day 
```{r}
stepsPerDayDT <- tapply(DT$steps, DT$date, sum)
qplot(stepsPerDayDT, xlab='Total Steps/ Day (NA replaced by Median/Interval)', ylab='Frequency Using Bin Width 500', binwidth=500, fill=I("blue"))
```

##### ... and Calculate and report the mean and median total number of steps taken per day. 
```{r}
stepsPerDayMeanDT <- mean(stepsPerDayDT)
stepsPerDayMedianDT <- median(stepsPerDayDT)
```
* Mean (NA replaced by Median/Interval): `r stepsPerDayMeanDT`
* Median (NA replaced by Median/Interval):  `r stepsPerDayMedianDT`

* Since the Mean is less than the Median and also because the method of replacing the NA values is by using the Median for each 5-minute interval, the new Steps Mean is expected to drop while the Median remains the same.
----

## Are there differences in activity patterns between weekdays and weekends?
##### 1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}
DT$dateType <-  ifelse(as.POSIXlt(DT$date)$wday %in% c(0,6), 'weekend', 'weekday')
```

##### 2. Make a panel plot containing a time series plot

```{r}
averageActivityDataDT <- aggregate(steps ~ interval + dateType, data=DT, mean)
ggplot(averageActivityDataDT, aes(interval, steps)) + 
    geom_line() + 
    facet_grid(dateType ~ .) +
    xlab("5-minute interval") + 
    ylab("avarage number of steps")
```

